---
title: "JFAITH_QBS103_final_project"
output: pdf_document
date: "2025-07-12"
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyr)
library(tidyverse)
library(ggplot2)
library(ggpubr)
```

# Comparing Biomarkers to Covid-19

## Data Wrangling

### Read in files

```{r}
gene_data <- read.csv("data/QBS103_GSE157103_genes.csv")
meta_data <- read.csv('data/QBS103_GSE157103_series_matrix-1.csv')
```

### Function To Create Gene Specific Dataframe

```{r}
#gene_name = 'ABCF1'
process_dataframe <- function(gene_name) {
  # Select gene
  gene <- gene_data[which(gene_data$X==gene_name), ]
  
  # Get all column names except X
  column_names <- names(gene)[-1]
  
  #Melt dataframe
  #referenced: https://tidyr.tidyverse.org/reference/pivot_longer.html
  gene_long <- pivot_longer(
                  gene, 
                  cols = all_of(column_names), 
                  names_to = "participant_id", 
                  values_to = "expression")
 
  gene_dataset <- inner_join(gene_long, meta_data, by="participant_id")
  # referenced: https://datascienceplus.com/merging-datasets-with-tidyverse/
   
   gene_dataset$HospitalDaysGroup <- cut(
      gene_dataset$hospital.free_days_post_45_day_followup,
      breaks = c(0, 10, 20, 30, 40, 50),
      labels = c('Under 10', '10-20', '20-30', '30-40', 'Over 40'),
      right = FALSE
    )
   return (gene_dataset)
}

```

## Visualizations

```{r}

#Custom Color Palette
colorPalette <- c('#004777', '#F05D5E', '#A8D0DB', '#136F63', '#FFC857', '#9DD9D2')

myTheme <- theme(
        panel.border = element_blank(), 
        panel.grid.major = element_line(colour="grey", linewidth = rel(.1)),
        panel.grid.minor= element_line(colour="grey", linewidth = rel(.1)),
        #panel.grid.minor = element_blank(),
        # Define my axis
        plot.title = element_text(colour = "black", hjust = .5, face='bold'),
        axis.line = element_line(colour = '#004777', linewidth = rel(1)),
        axis.title.x = element_text(color='black'),
        #axis.title.y = element_text(color='black'),
        axis.text = element_text(color='#004777'),
        # Set plot background
        panel.background = element_blank()
        )
```

Build a function to create the plots you made for Presentation 1, incorporating any feedback you received on your submission. Your functions should take the following input: (1) the name of the data frame, (2) a list of 1 or more gene names, (3) 1 continuous covariate, and (4) two categorical covariates (10 pts)

```{r results='asis'}

## documentation on working with .data[[]]
##https://dplyr.tidyverse.org/articles/programming.html

create_graphs <- function(
                          dataframe_name, 
                          gene_name, 
                          continuous_covariate, 
                          categorical_covariates
                          ) {
  
  #https://stackoverflow.com/questions/69338331/r-markdown-create-section-headers-in-r-and-print-r-output-in-same-section
  cat("\\newpage\n")
  cat(paste0("# Visuals for ", gene_name, "\n\n"))
  
  histogram_title <- paste(gene_name, "x Count of Participants")
  histogram <- ggplot(dataframe_name, aes(x = expression)) + 
              geom_histogram(binwidth=1, color="white", fill='#004777') + 
              labs(title=histogram_title, x = 'Gene Expression',y = 'Count') +
              myTheme
  
  print(histogram)
  
  cleaned_continuous_covariate <- gsub("[._]", " ", continuous_covariate)
  scatterplot_title <- paste("Gene Expression vs", cleaned_continuous_covariate)
  
  scatterplot <- ggplot(dataframe_name, aes(x = expression, y = .data[[continuous_covariate]])) + 
              geom_point() + 
              labs(title=scatterplot_title, x = 'Gene Expression',y = cleaned_continuous_covariate)+
              scale_color_manual(values = colorPalette)+
              myTheme
  print(scatterplot)
  
  boxplot_title <- paste(gene_name, 'vs', categorical_covariates[1])
  boxplot <- ggplot(
              dataframe_name,
              aes(
                x = .data[[categorical_covariates[1]]],
                y = expression,
                color = .data[[categorical_covariates[2]]]
              )) +
            geom_boxplot() +
            labs(
              title = boxplot_title,
              x = categorical_covariates[1],
              y = 'Gene Expression'
            ) +
            scale_color_manual(values = colorPalette) +
            myTheme
  print(boxplot)
}
```


# Displaying All Graphs
```{r results='asis'}
for (gene in c('ABCF1', 'AAGAB', 'ABAT')){
  gene_df <- process_dataframe(gene)
  create_graphs(gene_df, gene, 'hospital.free_days_post_45_day_followup', c('disease_status', 'sex'))
}
```

